The Error Estimate for Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald.

### 2012-04-20 longRange.hetero/publish/main.tex
add alltext
fix bugs

\documentclass[aps,pre,preprint]{revtex4}
% \documentclass[aps,pre,twocolumn]{revtex4-1}
% \documentclass[aps,jcp,groupedaddress,twocolumn,unsortedaddress]{revtex4}

@@

\title{The Error Estimate for Computing Electrostatic Interaction
  in the Inhomogeneous and Correlated Molecular Systems:
  for the Ewald Summation, SPME and Staggered Mesh Ewald.
}
\author{Han Wang}
\email{han.wang@fu-berlin.de}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

@@

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to try out the
most efficient parameter set in such a high dimensional parameter space.
The experiential choices may be either less accurate or waste
computational effort.
This problem can be solved by an
in-depth study of the error introduced by these algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one given a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid biomembrane
solving in water: the carbon chains of the lipids carry no charge, and
the charges are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  A lot of
research have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

@@

    q_j^2\vert\v K(\v r, \v r_j)\vert^2

@@

    q_jq_k \v K(\v r, \v r_j)\cdot\v K(\v r, \v r_k)

@@

Some minor contributions of the present work are:
\begin{itemize}
\item In a locally neutral charge system, the inhomogeneity
  error was proved to vanish. Therefore, the error
  is of good property, and no adaptive method nor
  force correction~\cite{wang2012} is necessary. Fortunately, most
  real molecular systems are locally neutral.
\item We explicitly gave the expression of the self-interaction term
  in the analytical differentiation, which was proved to dominant the
  force error in low density systems~\cite{cerutti2009staggered}.  The
  expression helped removing the self-interaction term at a very low
  computational cost.
\item The error estimate showed that the Staggered mesh Ewald
  (both ik- and analytical differentiation) are always more precise than
  the original SPME.
\item Unlike the original analytical differentiation
  that violates the Newton's third law, the staggered mesh analytical
  differentiation strictly preserves it.
\item We also proved the equivalence of the staggered mesh ik- and analytical
  differentiation, to the leading order of the force error.
  The staggered mesh analytical differentiation requires only one half
  FFTs as the ik-differentiation, so the former may be preferable
  for massive parallel simulation, due to the bottleneck of all-to-all
  communication required by the parallel FFT.
\end{itemize}

@@

\newpage
\appendix
\section{Error estimate of the smooth particle mesh
Ewald method}
\label{sec:appendix}

@@

\subsection{analytical differentiation}
Starting from Eqn.~\eqref{eqn:ana.error.force}, we have the error force kernel
of the analytical differentiation (notice the self-interacting term is removed):
\begin{align}
  \v K^{\textrm{ana}}_{\textrm{rec}}(\v r_i, \v r_j)
  =
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \{
  [A(\v m, \v r_i) + A(\v m,-\v r_j)\,]
  \v g(\v m) -
  2\v B(\v m,\v r_i) f(\v m)
  \}
  e^{-2\pi i\v m\cdot(\v r_i-\v r_j)}
\end{align}
By inserting Eqn.~\eqref{eqn:error-A} and \eqref{eqn:error-B}, we have
\begin{align}\nonumber
  \v K^{\textrm{ana}}_{\textrm{rec}}(\v r_i, \v r_j)
  =\,&
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  \big[
  \,(
  e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r_i} +
  e^{-2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r_j}
  - 2
  )
  \v g(\v m)
  \\  \label{eqn:ik.kernel.2}
  \,&
  - 4\pi i l K_\alpha\v a_\alpha^\ast e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r_i}
  f(\v m)
  \big]
  Z_{\alpha,l}(\v m)
  e^{-2\pi i\v m\cdot(\v r_i-\v r_j)}  
\end{align}
Following  the same idea as the error estimate of  ik-differentiation,
\begin{align}
  \langle\Delta\v F^{\textrm{ana}}_{\textrm{rec}}(\v r_i)\rangle
  \approx\,&
  -2q_i
  \big[
  \big(
  \sum_{\alpha}\sum_{l\neq 0}\v G_{\alpha,l}
  \big)
  \ast\rho_q\,
  \big] (\v r_i).
\end{align}
To calculate the RMS error, we start from
\begin{align}
  \vert\v K^{\textrm{ana}}_{\textrm{rec}}(\v r_i,\v r_j)\vert^2
  =\,&
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \sum_{
    \begin{subarray}{c}
      \vert n_{\alpha}\vert < K_\alpha/2\\
      \v n\neq 0
    \end{subarray}}
  \sum_{\alpha,\beta}
  \sum_{l_1,l_2}
  \\\nonumber
  &\,
  \big[
  \,(
  e^{2\pi i l_1 K_\alpha\v a^\ast_\alpha\cdot \v r_i} +
  e^{-2\pi i l_1 K_\alpha\v a^\ast_\alpha\cdot \v r_j}
  - 2
  )
  \v g(\v m)
  - 4\pi i l_1 K_\alpha\v a_\alpha^\ast e^{2\pi i l_1 K_\alpha\v a^\ast_\alpha\cdot \v r_i}
  f(\v m)
  \big] \\\nonumber
  \,&
  \times
  \big[
  \,(
  e^{2\pi i l_2 K_\beta\v a^\ast_\beta\cdot \v r_i} +
  e^{-2\pi i l_2 K_\beta\v a^\ast_\beta\cdot \v r_j}
  - 2
  )
  \v g(\v n)
  - 4\pi i l_2 K_\beta\v a_\beta^\ast e^{2\pi i l_2 K_\beta\v a^\ast_\beta\cdot \v r_i}
  f(\v n)
  \big]
  \\
  &\,
  \times
  Z_{\alpha,l_1}(\v m)Z_{\beta,l_2}(\v n)
  e^{-2\pi i(\v m-\v n)\cdot(\v r_i-\v r_j)}
\end{align}
Similar to the calculation of \eqref{eqn:ik.cal.tmp1}, one can eventually
prove the error estimate for the homogeneity RMS error:
\begin{align}\nonumber
  [\mathcal E^{\textrm{ana}}_{\textrm{homo}}(\v r_i)]^2
  = \,&
  \int
  \vert\v K^{\textrm{ana}}_{\textrm{rec}}(\v r_i,\v r')\vert^2
  \rho_{q^2}(\v r')\,\d d\v r' \\\nonumber
  \approx\,&  
  q_i^2
  \big[\,
  \big(
  \sum_{\alpha} \sum_{l\neq 0}
  \vert \v G_{\alpha,l}\vert^2
  \big)
  \ast \rho_{q^2}
  \,\big] (\v r_i) +
  q_i^2
  \big[\,
  \big(
  \sum_{\alpha} \sum_{l\neq 0}
  \vert \v G_{\alpha,l} +\v F_{\alpha,l}\vert^2
  \big)
  \ast \rho_{q^2}
  \,\big] (\v r_i) \\
  \,&+
  4q_i^2\,
  \big[
  \vert
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \vert^2
  \ast \rho_{q^2}\,
  \big] (\v r_i) 
\end{align}

@@

\newpage

>>>>

\documentclass[aps,pre,preprint,unsortedaddress]{revtex4}
% \documentclass[aps,pre,unsortedaddress]{revtex4-1}
%\documentclass[aps,pre,twocolumn]{revtex4-1}
%\documentclass[aps,jcp,groupedaddress,twocolumn,unsortedaddress]{revtex4}

@@

\title{The Error Estimate for Computing Electrostatic Interaction
  in the Inhomogeneous and Correlated Molecular Systems:
  for the Ewald Summation, SPME and Staggered Mesh Ewald.
}
\author{Han Wang}
\email{han.wang@fu-berlin.de}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Pingwen Zhang}
\affiliation{LMAM and School of Mathematical Sciences, Peking University}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

@@

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to try out the
most efficient parameter set in such a high dimensional parameter space.
The experiential choices may be either less accurate or waste
computational effort.
This problem can be solved by an
in-depth study of the error introduced by these algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2, stern2008mesh,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one given a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid biomembrane
solving in water: the carbon chains of the lipids carry no charge, and
the charges are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  A lot of
research have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

@@

    q_j^2\vert\v K(\v r_i, \v r_j)\vert^2

@@

    q_jq_k \v K(\v r_i, \v r_j)\cdot\v K(\v r_i, \v r_k)

@@

Some minor contributions of the present work are:
\begin{itemize}
\item In a locally neutral charge system, the inhomogeneity
  error was proved to vanish. Therefore, the error
  is of good property, and no adaptive method nor
  force correction~\cite{wang2012} is necessary. Fortunately, most
  real molecular systems are locally neutral.
\item We explicitly gave the expression of the self-interaction term
  in the analytical differentiation, which was proved to dominant the
  force error in low density systems~\cite{cerutti2009staggered}.  The
  expression helped removing the self-interaction term at a very low
  computational cost.
\item The error estimate showed that the Staggered mesh Ewald
  (both ik- and analytical differentiation) are always more precise than
  the original SPME.
\item Unlike the original analytical differentiation
  that violates the Newton's third law, the staggered mesh analytical
  differentiation preserves it to the leading order.
\item We also proved the equivalence of the staggered mesh ik- and analytical
  differentiation, to the leading order of the force error.
  The staggered mesh analytical differentiation requires only one half
  FFTs as the ik-differentiation, so the former may be preferable
  for massive parallel simulation, due to the bottleneck of all-to-all
  communication required by the parallel FFT.
\end{itemize}

@@

\appendix
\section{Error estimate of the smooth particle mesh
Ewald method}
\label{sec:appendix}

@@

\subsection{analytical differentiation}
Starting from Eqn.~\eqref{eqn:ana.error.force}, we have the error force kernel
of the analytical differentiation (notice the self-interacting term is removed):
\begin{align}
  \v K^{\textrm{ana}}_{\textrm{rec}}(\v r_i, \v r_j)
  =
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \{
  [A(\v m, \v r_i) + A(\v m,-\v r_j)\,]
  \v g(\v m) -
  2\v B(\v m,\v r_i) f(\v m)
  \}
  e^{-2\pi i\v m\cdot(\v r_i-\v r_j)}
\end{align}
By inserting Eqn.~\eqref{eqn:error-A} and \eqref{eqn:error-B}, we have
\begin{align}\nonumber
  \v K^{\textrm{ana}}_{\textrm{rec}}(\v r_i, \v r_j)
  =\,&
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  \big[
  \,(
  e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r_i} +
  e^{-2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r_j}
  - 2
  )
  \v g(\v m)
  \\ 
  \,&
  - 4\pi i l K_\alpha\v a_\alpha^\ast e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r_i}
  f(\v m)
  \big]
  Z_{\alpha,l}(\v m)
  e^{-2\pi i\v m\cdot(\v r_i-\v r_j)}  
\end{align}
Following  the same idea as the error estimate of  ik-differentiation,
\begin{align}
  \langle\Delta\v F^{\textrm{ana}}_{\textrm{rec}}(\v r_i)\rangle
  \approx\,&
  -2q_i
  \big[
  \big(
  \sum_{\alpha}\sum_{l\neq 0}\v G_{\alpha,l}
  \big)
  \ast\rho_q\,
  \big] (\v r_i).
\end{align}
To calculate the RMS error, we start from
\begin{align}
  \vert\v K^{\textrm{ana}}_{\textrm{rec}}(\v r_i,\v r_j)\vert^2
  =\,&
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  \sum_{
    \begin{subarray}{c}
      \vert n_{\alpha}\vert < K_\alpha/2\\
      \v n\neq 0
    \end{subarray}}
  \sum_{\alpha,\beta}
  \sum_{l_1,l_2}
  \\\nonumber
  &\,
  \big[
  \,(
  e^{2\pi i l_1 K_\alpha\v a^\ast_\alpha\cdot \v r_i} +
  e^{-2\pi i l_1 K_\alpha\v a^\ast_\alpha\cdot \v r_j}
  - 2
  )
  \v g(\v m)
  - 4\pi i l_1 K_\alpha\v a_\alpha^\ast e^{2\pi i l_1 K_\alpha\v a^\ast_\alpha\cdot \v r_i}
  f(\v m)
  \big] \\\nonumber
  \,&
  \times
  \big[
  \,(
  e^{2\pi i l_2 K_\beta\v a^\ast_\beta\cdot \v r_i} +
  e^{-2\pi i l_2 K_\beta\v a^\ast_\beta\cdot \v r_j}
  - 2
  )
  \v g(\v n)
  - 4\pi i l_2 K_\beta\v a_\beta^\ast e^{2\pi i l_2 K_\beta\v a^\ast_\beta\cdot \v r_i}
  f(\v n)
  \big]
  \\
  &\,
  \times
  Z_{\alpha,l_1}(\v m)Z_{\beta,l_2}(\v n)
  e^{-2\pi i(\v m-\v n)\cdot(\v r_i-\v r_j)}
\end{align}
Similar to the calculation of \eqref{eqn:ik.cal.tmp1}, one can eventually
prove the error estimate for the homogeneity RMS error:
\begin{align}\nonumber
  [\mathcal E^{\textrm{ana}}_{\textrm{homo}}(\v r_i)]^2
  = \,&
  \int
  \vert\v K^{\textrm{ana}}_{\textrm{rec}}(\v r_i,\v r')\vert^2
  \rho_{q^2}(\v r')\,\d d\v r' \\\nonumber
  \approx\,&  
  q_i^2
  \big[\,
  \big(
  \sum_{\alpha} \sum_{l\neq 0}
  \vert \v G_{\alpha,l}\vert^2
  \big)
  \ast \rho_{q^2}
  \,\big] (\v r_i) +
  q_i^2
  \big[\,
  \big(
  \sum_{\alpha} \sum_{l\neq 0}
  \vert \v G_{\alpha,l} +\v F_{\alpha,l}\vert^2
  \big)
  \ast \rho_{q^2}
  \,\big] (\v r_i) \\
  \,&+
  4q_i^2\,
  \big[
  \vert
  \sum_{\alpha} \sum_{l\neq 0}  
  \v G_{\alpha,l}
  \vert^2
  \ast \rho_{q^2}\,
  \big] (\v r_i) 
\end{align}

@@
